---
title: Advanced HRF Modeling and Design
author: Bradley R. Buchsbaum
date: '`r Sys.Date()`'
output:
  albersdown::albers_vignette:
    family: red
    preset: interaction
    toc: yes
    toc_depth: 2.0

vignette: |
  %\VignetteIndexEntry{Advanced HRF Modeling and Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Sharper website figures; keep the CRAN vignette archive compact.
  fig.retina = if (identical(Sys.getenv("IN_PKGDOWN"), "true")) 2 else 1,
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
# CRAN builds: skip dark-mode figure twins to keep the source package small;
# the pkgdown site (IN_PKGDOWN = "true") keeps them.
if (!identical(Sys.getenv("IN_PKGDOWN"), "true")) {
  options(albersdown.dark_figures = FALSE)
}
library(fmrihrf)
library(dplyr)
library(ggplot2)
library(tidyr)
library(Matrix)
```

## Introduction

This vignette explores advanced features of `fmrihrf` for systematic HRF modeling, regularization, and experimental design. We'll cover four key functions that extend the basic HRF framework:

- **`hrf_library()`**: Creating systematic collections of HRF variants
- **`reconstruction_matrix()`**: Converting basis coefficients back to HRF shapes
- **`regressor_set()`**: Managing multi-condition experimental designs
- **`regressor_design()`**: Building design matrices for complex experimental blocks

These tools are essential for advanced fMRI modeling where you need flexibility in HRF specification, robust estimation with limited data, or complex experimental designs.

## HRF Libraries: Systematic Parameter Exploration

The `hrf_library()` function creates collections of HRF variants by systematically varying parameters. This is useful for exploring how different HRF assumptions affect your model or for building data-driven HRF basis sets.

### Example 1: Library of Gamma HRFs

Let's create a library of gamma HRFs with different shape and rate parameters:

```{r gamma_library, fig.height=6, fig.alt="Nine gamma HRFs grouped into panels by shape (4, 6, 8), with rate (0.8, 1, 1.2) distinguished by color. Higher rates peak earlier and higher; all panels share the same axes."}
# Define parameter grid for gamma HRFs
gamma_params <- expand.grid(
  shape = c(4, 6, 8),
  rate = c(0.8, 1.0, 1.2)
)
print(gamma_params)

# Create a generator function for gamma HRFs
make_gamma_hrf <- function(shape, rate) {
  gen_hrf(hrf_gamma, shape = shape, rate = rate, name = paste0("Gamma_", shape, "_", rate))
}

# Create HRF library
gamma_lib <- hrf_library(make_gamma_hrf, gamma_params)
print(gamma_lib)
nbasis(gamma_lib) # 9 HRFs total (3 x 3 grid)

# Evaluate and visualize
time_points <- seq(0, 20, by = 0.1)
gamma_responses <- gamma_lib(time_points)

# Convert to long format for plotting
gamma_df <- as.data.frame(gamma_responses)
names(gamma_df) <- with(gamma_params, paste(shape, rate, sep = " / "))
gamma_df$Time <- time_points

gamma_long <- pivot_longer(gamma_df, -Time, names_to = "Parameters", values_to = "Response")
gamma_long <- gamma_long %>%
  separate(Parameters, into = c("Shape", "Rate"), sep = " / ", remove = FALSE) %>%
  mutate(Shape = factor(Shape, levels = c("4", "6", "8")),
         Rate = factor(Rate, levels = c("0.8", "1", "1.2")))

# One panel per shape; rate (an ordered parameter) uses the ordered palette
ggplot(gamma_long, aes(x = Time, y = Response, color = Rate)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~Shape, ncol = 1, labeller = label_both) +
  scale_colour_hrf("ordered") +
  scale_y_continuous(breaks = c(0, 0.2)) +
  labs(title = "Gamma HRF library", subtitle = "Peak time = (shape - 1) / rate",
       x = "Time (s)", y = "Response", color = "Rate") +
  theme(legend.position = "bottom", legend.justification = "left",
        strip.text = element_text(hjust = 0), plot.title.position = "plot")
```

### Example 2: Library of Lagged SPM HRFs

Here's how to create a library of the SPM canonical HRF with different temporal lags:

```{r spm_lag_library, fig.alt="Seven SPM canonical HRFs lagged from -2 to +4 seconds in 1 second steps, coloured from violet (earliest) to ochre (latest). The shapes are identical and evenly spaced in time."}
# Parameter grid for temporal lags
lag_params <- data.frame(lag = seq(-2, 4, by = 1))
print(lag_params)

# Create library using a helper function that applies lag_hrf
create_lagged_spm <- function(lag) {
  lag_hrf(HRF_SPMG1, lag = lag)
}

spm_lag_lib <- hrf_library(create_lagged_spm, lag_params)
print(spm_lag_lib)

# The library is a basis set: one column per lag
plot_hrfs(spm_lag_lib, time = time_points,
          labels = sprintf("%+d s", lag_params$lag), palette = "ordered",
          title = "Library of lagged SPM canonical HRFs",
          subtitle = "Lags from -2 to +4 s")
```


## Reconstruction Matrices: From Coefficients to HRF Shapes

The reconstruction process converts a set of basis coefficients into a continuous HRF shape. Understanding this transformation is key to interpreting estimated HRFs from fMRI analyses.

### How Reconstruction Works

A flexible basis set describes an HRF by a vector of weights, one per basis
function. The reconstruction matrix holds the basis functions evaluated on a
time grid (one column per function), so multiplying it by the weights gives
the HRF on that grid.

```{r reconstruction_demo}
# Ten cubic B-splines with 24-second support
basis_set <- hrf_bspline_generator(nbasis = 10, span = 24)
eval_times <- seq(0, 24, by = 0.1)

# The reconstruction matrix: each column is a basis function evaluated at time points
recon_matrix <- reconstruction_matrix(basis_set, eval_times)
dim(recon_matrix)
```

Every function is zero at 0 s and at 24 s, so any weighted sum starts and
ends at baseline.

```{r reconstruction_basis, fig.alt="Ten cubic B-spline basis functions on 0 to 24 seconds, labelled B1 to B10 at their peaks and coloured from violet (early) to ochre (late)."}
plot_hrfs(basis_set, time = eval_times,
          title = "Cubic B-spline basis, N = 10",
          subtitle = "24 s span; each function covers part of it")
```

In an analysis the weights come from the GLM. Here we obtain them by
least-squares fits of three target responses: the SPM canonical HRF, the same
response delayed by 3 s, and the response to a sustained 6-second event. The
basis is the same each time; only the weights change.

```{r reconstruction_fit}
targets <- list(
  "Canonical" = HRF_SPMG1,
  "Delayed" = lag_hrf(HRF_SPMG1, 3),      # canonical, 3 s later
  "Sustained" = block_hrf(HRF_SPMG1, width = 6, normalize = TRUE)  # 6 s event
)
fits <- lapply(targets, function(h) {
  y <- h(eval_times)
  y <- y / max(y)                       # unit peak
  w <- qr.solve(recon_matrix, y)        # least-squares weights
  list(weights = w, fitted = drop(recon_matrix %*% w), target = y)
})

# Fit quality (R^2) for each target
sapply(fits, function(f) {
  1 - sum((f$target - f$fitted)^2) / sum((f$target - mean(f$target))^2)
})

canonical_coefs <- fits[["Canonical"]]$weights
round(canonical_coefs, 2)
```

### Building an HRF from Weighted Basis Functions

Each panel shows the weighted basis functions (thin grey lines), their sum,
the reconstructed HRF (thick coloured line), and the target it was fitted to
(dashed). The canonical response is carried mainly by B3, with help from B2
and B4; its undershoot comes from the small negative weights on B6–B8. The fits are not
perfect: the small dip in the first second of the canonical fit, and the
wiggle before the delayed rise, are fitting errors, not modelled features.
Because every basis function is zero at 24 s, the fits also return to zero
there even where a target is still slightly below baseline (most visibly the
sustained response's undershoot).

```{r reconstruction_components, fig.height = 4.4, fig.alt="Three panels for the canonical, delayed and sustained targets. In each, thin grey curves are the ten basis functions scaled by their weights, a thick coloured curve is their sum, and a dashed curve is the target. The canonical sum peaks at 5 seconds, the delayed at 8 seconds, and the sustained around 8 seconds with a broader peak."}
component_df <- do.call(rbind, lapply(names(fits), function(nm) {
  w <- fits[[nm]]$weights
  data.frame(
    Time = rep(eval_times, length(w)),
    Value = as.vector(sweep(recon_matrix, 2, w, `*`)),
    Basis = factor(rep(paste0("B", seq_along(w)), each = length(eval_times)),
                   levels = paste0("B", seq_along(w))),
    Target = nm
  )
}))
sum_df <- do.call(rbind, lapply(names(fits), function(nm) {
  data.frame(Time = eval_times, Value = fits[[nm]]$fitted,
             Target_value = fits[[nm]]$target, Target = nm)
}))
component_df$Target <- factor(component_df$Target, levels = names(fits))
sum_df$Target <- factor(sum_df$Target, levels = names(fits))

ggplot(component_df, aes(Time, Value)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_line(aes(group = Basis), colour = "grey60", linewidth = 0.4) +
  geom_line(data = sum_df, aes(colour = Target), linewidth = 1.1) +
  geom_line(data = sum_df, aes(y = Target_value), colour = "grey15",
            linewidth = 0.5, linetype = "22") +
  facet_wrap(~Target, nrow = 1) +
  scale_colour_hrf() +
  scale_x_continuous(breaks = c(0, 10, 20)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(title = "Same basis, different weights",
       subtitle = "Grey: weighted basis functions\nColour: their sum. Dashed: target",
       x = "Time (s)", y = "Response / peak") +
  theme(legend.position = "none", strip.text = element_text(hjust = 0),
        plot.title.position = "plot")
```

The weights themselves summarize the shape. For the delayed response the
largest weight moves from B3 to B4 (and B2 turns slightly negative to delay
the rise); a sustained event spreads it over B3 to B5:

```{r reconstruction_coefficients, fig.height = 5.6, fig.alt="Lollipop charts of the ten basis weights for each target, coloured like the fitted curves above. The canonical fit puts its largest weight on B3, the delayed fit on B4, and the sustained fit on B3 to B5; later weights are small and mostly negative."}
coef_df <- do.call(rbind, lapply(names(fits), function(nm) {
  w <- fits[[nm]]$weights
  data.frame(Basis = factor(paste0("B", seq_along(w)), levels = paste0("B", seq_along(w))),
             Weight = w, Target = nm)
}))
coef_df$Target <- factor(coef_df$Target, levels = names(fits))

coef_df$Index <- as.integer(coef_df$Basis)
ggplot(coef_df, aes(Index, Weight, colour = Target)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_segment(aes(xend = Index, y = 0, yend = Weight), linewidth = 0.9) +
  geom_point(size = 2.2) +
  facet_wrap(~Target, ncol = 1) +
  scale_colour_hrf() +
  scale_x_continuous(breaks = 1:10, minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(title = "Basis weights for each target", x = "Basis function (B1-B10)",
       y = "Weight") +
  theme(legend.position = "none", strip.text = element_text(hjust = 0),
        plot.title.position = "plot")
```

## Regressor Sets: Multi-Condition Experimental Designs

The `regressor_set()` function simplifies creating regressors for multi-condition experiments where each condition shares the same HRF but has different event timings.

```{r regressor_set_demo}
# Simulate a 3-condition experiment
set.seed(123)
n_events_per_condition <- 8
total_duration <- 240  # 4 minutes

# Generate random onsets for each condition
condition_A_onsets <- sort(runif(n_events_per_condition, 0, total_duration))
condition_B_onsets <- sort(runif(n_events_per_condition, 0, total_duration))
condition_C_onsets <- sort(runif(n_events_per_condition, 0, total_duration))

# Combine all onsets and create factor
all_onsets <- c(condition_A_onsets, condition_B_onsets, condition_C_onsets)
conditions <- factor(rep(c("TaskA", "TaskB", "TaskC"), each = n_events_per_condition))

# Create regressor set
reg_set <- regressor_set(onsets = all_onsets, fac = conditions, hrf = HRF_SPMG1)
print(reg_set)

# Evaluate at scan times (TR = 2s)
TR <- 2
scan_times <- seq(0, total_duration, by = TR)
design_matrix <- evaluate(reg_set, scan_times)

print(dim(design_matrix)) # Time points x 3 conditions
```

`plot_regressors()` accepts a regressor set directly. With `layout = "stack"`,
each condition gets its own panel and its own events are marked below its
curve. `scales = "fixed"` puts all panels on one y axis, so response sizes can
be compared across conditions:

```{r regressor_set_plot, fig.height = 5, fig.alt="Three stacked panels, one per condition, each showing that condition's predicted BOLD response over 240 seconds with its eight event onsets marked below. Closely spaced events produce larger, merged responses."}
plot_regressors(reg_set, grid = seq(0, total_duration, by = 0.1),
                layout = "stack", scales = "fixed",
                title = "Multi-condition design",
                subtitle = "8 random onsets per condition")
```

## Regressor Design: Complex Block Designs

For more complex experimental designs with multiple blocks or runs, `regressor_design()` provides a higher-level interface that handles block-relative timing and creates design matrices directly.

```{r regressor_design_demo, fig.height = 4.4, fig.alt="Faces and Houses responses in two stacked panels across two 240-second blocks, with points at the 2 second scan samples. A dashed line marks the block boundary at 240 seconds; each block has its own event schedule."}
# Create a sampling frame for 2 blocks of 120 scans (240 s) each
sframe <- sampling_frame(
  blocklens = c(120, 120),  # Two 4-minute blocks (120 scans each at TR = 2s)
  TR = 2                    # 2-second TR
)
print(sframe)

# Generate block-relative event onsets
# Block 1: Faces at 10, 50, 90; Houses at 30, 70 seconds
# Block 2: Faces at 15, 55, 95; Houses at 35, 75 seconds  
block_onsets <- c(10, 30, 50, 70, 90, 15, 35, 55, 75, 95)
block_ids <- c(rep(1, 5), rep(2, 5))
event_conditions <- factor(c("Faces", "Houses", "Faces", "Houses", "Faces", 
                            "Faces", "Houses", "Faces", "Houses", "Faces"))

# Create design matrix using regressor_design
design_mat <- regressor_design(
  onsets = block_onsets,
  fac = event_conditions,
  block = block_ids,
  sframe = sframe,
  hrf = HRF_SPMG1
)

print(dim(design_mat)) # Total time points across both blocks x 2 conditions

# The same design on a 0.1 s grid shows the continuous responses; the design
# matrix itself is those responses sampled once per scan (TR = 2 s).
sframe_fine <- sampling_frame(blocklens = c(2400, 2400), TR = 0.1, precision = 0.05)
design_fine <- regressor_design(onsets = block_onsets, fac = event_conditions,
                                block = block_ids, sframe = sframe_fine, hrf = HRF_SPMG1)
fine_df <- data.frame(Time = rep(samples(sframe_fine, global = TRUE), 2),
                      Response = as.vector(design_fine),
                      Condition = rep(c("Faces", "Houses"), each = nrow(design_fine)))
scan_df <- data.frame(Time = rep(samples(sframe, global = TRUE), 2),
                      Response = as.vector(design_mat),
                      Condition = rep(c("Faces", "Houses"), each = nrow(design_mat)))
scan_df <- scan_df[abs(scan_df$Response) > 0.01, ]   # samples during responses
# Highest scan sample in each block, labelled on the Faces panel
peak_df <- do.call(rbind, lapply(split(scan_df[scan_df$Condition == "Faces", ],
                                       scan_df$Time[scan_df$Condition == "Faces"] > 240),
                                 function(d) d[which.max(d$Response), ]))

ggplot(fine_df, aes(Time, Response, colour = Condition)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_vline(xintercept = 240, colour = "grey45", linetype = "dashed") +
  geom_line(linewidth = 0.7) +
  geom_point(data = scan_df, size = 0.9) +
  geom_text(data = peak_df, aes(label = sprintf("max %.3f", Response)),
            hjust = -0.15, vjust = -0.4, size = 3, show.legend = FALSE) +
  facet_wrap(~Condition, ncol = 1) +
  scale_colour_hrf() +
  scale_y_continuous(breaks = c(0, 0.15), expand = expansion(mult = c(0.05, 0.35))) +
  labs(title = "Two-block design",
       subtitle = "Points: scan samples (TR = 2 s)\nDashed line: start of block 2",
       x = "Time (s)", y = "Predicted BOLD response") +
  theme(legend.position = "none", strip.text = element_text(hjust = 0),
        plot.title.position = "plot")

# Show global vs block-relative timing
timing_df <- data.frame(
  Block = block_ids,
  Block_Relative_Onset = block_onsets,
  Global_Onset = global_onsets(sframe, block_onsets, block_ids),
  Condition = event_conditions
)

print(timing_df)
```

The curves are the same in both blocks, but the scan samples (points) are not.
Scans are taken at odd seconds (the middle of each 2-second TR). Block 1's
events (10, 30, ... s) peak 5 s later, at odd seconds, so a scan lands on each
peak (0.175). Block 2 starts at 240 s, so its events fall at 255, 275, ... s
and peak at even seconds, between two scans; the highest sampled value is
0.160. This is a property of the design
matrix sampled at TR = 2 s, not a difference between blocks; it is one reason
to jitter onsets relative to the scan grid.
